Data Encryption Standard Implementation 4003-482-01 RIT \\ Team BitShift

### 2010-05-06 paper/paper.tex
Modify color
Paper finalizations.

\title{ Data Encryption Standard Implementation 4003-482-01 RIT }

@@

\definecolor{light-gray}{gray}{0.95}

@@

  {DES} is the 64-bit block cipher that was once used to encrypt all US Government non-classified documents.
  It was developed as a collaboration between NSA and IBM in the 1970's and was eventually selected
  by the National Institute of Standards and Technologies and was  standardised on January 15, 1977 as FIPS-46.
  As mentioned before the cipher has a 64-bit data block and uses a 64-bit key block, it should be noted though that
  only 56 bits of the key are ever actually used. The algorithm is based around sixteen rounds each with it's own
  shortened subkey (48 bits) based on the original key block. At the beginning of the cipher the data bock is permuted
  and is then inversely permuted

@@

  \subsection{Example 1}
  \textit{Hello}
  \\\indent Original:
  \begin{verbatim}
     48 65 6C 6C 6F 0A
  \end{verbatim}
  \indent Encrypted:
  \begin{verbatim}
     94 93 81 0E 62 47 30 86
  \end{verbatim}

  \subsection{Example 2}
  \textit{World}
  \\\indent Original:
  \begin{verbatim}
     57 6F 72 6C 64 0A
  \end{verbatim}
  \indent Encrypted:
  \begin{verbatim}
     24 0B A5 B2 CB C1 F2 8B
  \end{verbatim}

  \subsection{Example 3}
  \textit{Hello World}
  \\\indent Original:
  \begin{verbatim}
     48 65 6C 6C 6F 20 57 6F
     72 6C 64 0A
  \end{verbatim}
  \indent Encripted:
  \begin{verbatim}
     E6 74 22 8A AC 1E 85 ED
     B7 3C 26 3C 4B B4 F1 77
  \end{verbatim}

@@

\subsection{Example 4}
\textit{The quick brown fox jumps over the lazy dog}
\\\indent Original:
\begin{verbatim}
   54 68 65 20 71 75 69 63
   6B 20 62 72 6F 77 6E 20
   66 6F 78 20 6A 75 6D 70
   73 20 6F 76 65 72 20 74
   68 65 20 6C 61 7A 79 20
   64 6F 67 0A
\end{verbatim}
\indent Encrypted:
\begin{verbatim}
   09 D1 3F 78 9D C7 66 B7
   46 52 A9 9F 57 92 38 65
   8B 40 39 1C 21 9E 77 D7
   9B 84 62 AF E9 22 25 1D
   96 61 A7 22 09 97 2C D2
   16 F5 D8 38 E4 FF C6 86
\end{verbatim}

@@

The general ideology of this first iteration is that we want the code to be clean and correct.
We are not putting too much thought into optimization and efficiency. We are very familiar
with the old Donal Knuth quote:

\begin{quote}
   "We should forget about small efficiencies, say about 97\% of the time: premature optimization is the root of all evil"
\end{quote}

During the duration of the project we attempted to work in a {TDD} model.
All major functions and actions have associated unit-tests which we derived from various peices of
literature on DES. These unit test's were then hooked up to run using the Google Test Framework \cite{gtest}.

\section{Original Runtime Measurements}
\paragraph{}Each test consisted of 1,000,000 encryptions of the message 123456789ABCDEF and key 133457799BBCDFF1. They were performed on a 1.8 Ghz Core2Duo computer running Ubuntu 10.04.
\\Test 1:
\\ \indent 10.899 seconds
\\Test 2:
\\ \indent 10.867 seconds
\\Test 3:
\\ \indent 10.870 seconds
\\\\Average:
\\ \indent 10.879 seconds
\begin{verbatim}
  % of   # of      # of
  time  Seconds    Calls      Name
 40.93   18.00   51,000,000   DES::mixer()
 22.88   10.06   16,000,000   DES::f()
 20.17    8.87    1,000,000   DES::keyschedule()
  8.20    3.60    1,000,000   DES::algorithm()
  5.59    2.46  512,000,000   DES::get()
  0.61    0.27   32,000,000   std::swap<T>()
  0.53    0.23    1,000,000   std::swap_ranges<T,G>()
  0.41    0.18   32,000,000   std::__iter_swap<T>::iter_swap<G,E>()
  0.28    0.12   32,000,000   std::iter_swap<T,G>()
  0.20    0.09                DES::on()
  0.11    0.05    1,000,000   DES::DES()
  0.02    0.01    1,000,000   DES::encrypt()
  0.02    0.01    1,000,000   DES::~DES()
  0.02    0.01                DES::main
\end{verbatim}
\section{Original Runtime Analysis and Design Changes}
\indent With a total run time of 10.884 seconds, this means each encryption took only .01 milliseconds.  To find improvements, we look at our profiling.  The most time was spent in our mixer function.  To reduce this we decided to 'unravel' the loop.  To do this we would have had to make a unique method for each of the different mixing procedures.  However instead we decided to get rid of it all together by putting the unraveled contents wherever mixer() was called from.
\\ \indent The next worst function was the f (feistel) function.  One of the first things we noticed could be done for improvment was to remove the first for-loop. First it mixes the input, then runs through it again doing xor with the key. We combined the two steps into the already unraveled and inlined mix procedure. Also, instead of using our get method to retrieve bit values, we copied the code into the function and hard coded the values that were needed for each use.
\\ \indent After this we analyzed the keyschedule algorithm, as this was \#3 in terms of time spent executing.  We had already internalized the mixer function.  The innermost for loop was unraveld to remove time spent computing the the counter and comparisons.  Also, we turned the for-loop outside of it into a do-while loop.  This made modest savings since in cases where it only loops once, since there are no extra jumps.  At the end of the function was one of the unraveled mix algorithms.  We combined this with the for loop afterwards so that we weren't running through each element twice.

\section{Revised Runtime Measurements}
\paragraph{}Each test consisted of 1,000,000 encryptions of the message 123456789ABCDEF and key 133457799BBCDFF1. They were performed on a 1.8 Ghz Core2Duo computer running Ubuntu 10.04.
\\Test 1:
\\ \indent 5.960 seconds
\\Test 2:
\\ \indent 6.012 seconds
\\Test 3:
\\ \indent 5.903 seconds
\\\\Average:
\\ \indent 5.958 seconds
\begin{verbatim}
  % of   # of      # of
  Time  Seconds    Calls      Name
 42.90    6.10   16,000,000   DES::f()
 24.40    3.47    1,000,000   DES::keyschedule()
 24.19    3.44    1,000,000   DES::algorithm()
  2.04    0.29    1,000,000   DES::inverseInitPermutation()
  2.04    0.29    1,000,000   std::swap_ranges<T,T>()
  1.30    0.18   32,000,000   std::iter_swap<T,T>()
  1.02    0.14   32,000,000   std::__iter_swap<G>::iter_swap<T,T>()
  0.91    0.13    1,000,000   DES::initPermutation()
  0.70    0.10   32,000,000   std::swap<unsigned char>()
  0.21    0.03    1,000,000   DES::~DES()
  0.14    0.02    1,000,000   DES::DES()
  0.07    0.01    1,000,000   DES::encrypt()
  0.07    0.01                main
\end{verbatim}

\section{Revised Runtime Analysis}
  One of the major changes was the f (feistel) function. It went from 23\% to 43\% of the total runtime.  This change is largely because we removed the mixer function. Mixer accounted for 41\% of the runtime, 63\% of mixer's calls were from feistel.  From this we can deduce that 26\% out of the 41\% was directly due to f, and we could say that f really caused 23\% + 26\% = 49\% of the runtime.  From this we can see that the f function's overall contribution to execution time went down without removing any of its duties. 
\section{Developers manual}

@@

... test output ...
\end{verbatim}
}

\section{Users Manual}

\noindent In order to encrypt a file using our implementation the user must run the executable \textbf{encrypt\_opt}.
\begin{verbatim}

  Usage: ./encrypt_opt <infile> <outfile> <key>

    <infile>  - The file you wish to encrypt
    <outfile> - The resulting encrypted file
    <key>     - 8 character string for your key

\end{verbatim}

\noindent The user has a similar choice for decrypting, both an optimized and original implementation exist.
\begin{verbatim}

  Usage: ./decrypt_opt <infile> <outfile> <key>

    <infile>  - The file you wish to decrypt
    <outfile> - The resulting decrypted file
    <key>     - 8 character string for your key
    <key>     - 8 character string for your key

\end{verbatim}

@@

  \textbf{Brian:}
  \begin{itemize}
    \item Unit test setup
    \item Build system
    \item Gathering of all constants and permutation blocks.
    \item Key schedule
    \item Main DES algorithm
  \end{itemize}

  \textbf{Sam:}
  \begin{itemize}
    \item Bit operators and converters
    \item Feistel function
    \item Implementation API design
    \item Mix function to handle mixing
    \item Block mode implementation
  \end{itemize}

@@

  \bibitem{gtest}{ \href{ http://code.google.com/p/googletest/ }{ Google Unit Testing Framework } }
  \bibitem{ spec  }{ \href{ http://csrc.nist.gov/publications/fips/fips46-3/fips46-3.pdf }{ Data Encryption Standard Specification } }
  \bibitem{ wiki  }{ \href{ http://en.wikipedia.org/wiki/Data_Encryption_Standard }{ Wikipedia: Data Encryption Standard } }
  \bibitem{ block }{ \href{ http://en.wikipedia.org/wiki/Block_cipher_modes_of_operation }{ Wikipedia: Block Cipher Modes } }
  \bibitem{ desge }{ \href{ http://www.comms.scitech.sussex.ac.uk/fft/crypto/des_algorithm_details.txt}{ How to implement the Data Encryption Standard (DES) } }
  \bibitem{ desg2 }{ \href{ http://orlingrabbe.com/des.htm }{ DES Illustrated } }
  \bibitem{ gprof }{ \href{ http://linux.die.net/man/1/gprof }{ GProf Manual Page } }
  \bibitem{ g++ }{ \href{ http://linux.die.net/man/1/g++ }{ GCC Manual Page } }

>>>>

\title{ Data Encryption Standard Implementation 4003-482-01 RIT \\ Team BitShift }

@@

% \raggedcolumns

\definecolor{light-gray}{gray}{0.50}

@@

\indent DES\cite{spec} is the 64-bit block cipher that was once used to encrypt all US Government 
  non-classified documents.  It was developed primarily by IBM with collaboration with the NSA
  in the 1970's and was eventually selected by the National Institute of Standards   
  and Technologies (NIST) as the first standarized block cipher.
  It was standardised on January 15, 1977 as FIPS-46. As mentioned 
  before the cipher has a 64-bit data block and uses a 64-bit key block, it should be 
  noted though that only 56 bits of the key are ever actually used. The algorithm is 
  based around sixteen rounds each with it's own shortened subkey (48 bits) based on 
  the original key block. DES has since been broken, one attack used a large distributed computing network
  to crack a DES key in a mere 22 hours of compute time \cite{deepcrack}. Many cryptographers cite
  it's primarily weekness of DES is it's 56-bit key, being far too small. 
  As of July 2005 NIST has withdrawn DES (FIPS-46) as a standardized
  cryptographic primative for govenment use\cite{withdrawn}. The cipher served the the US government
  and the world at large for around twenty eight years of secure service.

@@

  \subsection{Example 1}

  \textit{Hello} \\
\noindent Original:
  \begin{verbatim}
     48 65 6C 6C 6F 0A
  \end{verbatim}
  \indent Encrypted:
  \begin{verbatim}
     94 93 81 0E 62 47 30 86
  \end{verbatim}

  \subsection{Example 2}

  \textit{World} \\
\noindent Original:
  \begin{verbatim}
     57 6F 72 6C 64 0A
  \end{verbatim}
  \indent Encrypted:
  \begin{verbatim}
     24 0B A5 B2 CB C1 F2 8B
  \end{verbatim}

  \subsection{Example 3}

  \textit{Hello World} \\
\noindent Original:
  \begin{verbatim}
     48 65 6C 6C 6F 20 57 6F
     72 6C 64 0A
  \end{verbatim}
  \indent Encrypted:
  \begin{verbatim}
     E6 74 22 8A AC 1E 85 ED
     B7 3C 26 3C 4B B4 F1 77
  \end{verbatim}

@@

\subsection{Example 4}

\textit{The quick brown fox jumps over the lazy dog} \\

\noindent Original:
\begin{verbatim}
   54 68 65 20 71 75 69 63
   6B 20 62 72 6F 77 6E 20
   66 6F 78 20 6A 75 6D 70
   73 20 6F 76 65 72 20 74
   68 65 20 6C 61 7A 79 20
   64 6F 67 0A
\end{verbatim}
\indent Encrypted:
\begin{verbatim}
   09 D1 3F 78 9D C7 66 B7
   46 52 A9 9F 57 92 38 65
   8B 40 39 1C 21 9E 77 D7
   9B 84 62 AF E9 22 25 1D
   96 61 A7 22 09 97 2C D2
   16 F5 D8 38 E4 FF C6 86
\end{verbatim}

@@

During the duration of the project we attempted to work in a Test Driven Development (TDD) model.
All major functions and actions have associated unit-tests which we derived from various peices of
literature on DES. These unit test's were then hooked up to run using the Google Test Framework \cite{gtest}.
The general ideology of this first iteration is that we want the code to be clean and correct.
We are not putting too much thought into optimization and efficiency. We are very familiar
with the old Donal Knuth quote:

{
\textbf{
\large{
\begin{quote}
   ``We should forget about small efficiencies, say about 97\% of the time: premature optimization is the root of all evil.''
\end{quote}
}
}
}

\section{Original Runtime Measurements}
\paragraph{}Each test consisted of 1,000,000 encryptions of the message 123456789ABCDEF and key 133457799BBCDFF1. They were performed on a 1.8 Ghz Core2Duo computer running Ubuntu 10.04.

\begin{multicols}{4}
Test 1: 10.899 sec

Test 2: 10.867 sec

Test 3: 10.870 sec

Average: $ 10.879 $ sec
\end{multicols}
\begin{verbatim}
  % of   # of      # of
  time  Seconds    Calls      Name
 40.93   18.00   51,000,000   DES::mixer()
 22.88   10.06   16,000,000   DES::f()
 20.17    8.87    1,000,000   DES::keyschedule()
  8.20    3.60    1,000,000   DES::algorithm()
  5.59    2.46  512,000,000   DES::get()
  0.61    0.27   32,000,000   std::swap<T>()
  0.53    0.23    1,000,000   std::swap_ranges<T,G>()
  0.41    0.18   32,000,000   std::__iter_swap<T>::iter_swap<G,E>()
  0.28    0.12   32,000,000   std::iter_swap<T,G>()
  0.20    0.09                DES::on()
  0.11    0.05    1,000,000   DES::DES()
  0.02    0.01    1,000,000   DES::encrypt()
  0.02    0.01    1,000,000   DES::~DES()
  0.02    0.01                DES::main
\end{verbatim}
\section{Original Runtime Analysis and Design Changes}
\indent With a total run time of 10.884 seconds, this means each encryption took only .01 milliseconds.  To find improvements, we look at our profiling.  The most time was spent in our mixer function.  To reduce this we decided to 'unravel' the loop.  To do this we would have had to make a unique method for each of the different mixing procedures.  However instead we decided to get rid of it all together by putting the unraveled contents wherever mixer() was called from.
\\ \indent The next worst function was the f (feistel) function.  One of the first things we noticed could be done for improvment was to remove the first for-loop. First it mixes the input, then runs through it again doing xor with the key. We combined the two steps into the already unraveled and inlined mix procedure. Also, instead of using our get method to retrieve bit values, we copied the code into the function and hard coded the values that were needed for each use.
\\ \indent After this we analyzed the keyschedule algorithm, as this was \#3 in terms of time spent executing.  We had already internalized the mixer function.  The innermost for loop was unraveld to remove time spent computing the the counter and comparisons.  Also, we turned the for-loop outside of it into a do-while loop.  This made modest savings since in cases where it only loops once, since there are no extra jumps.  At the end of the function was one of the unraveled mix algorithms.  We combined this with the for loop afterwards so that we weren't running through each element twice.

\section{Revised Runtime Measurements}
\paragraph{}Each test consisted of 1,000,000 encryptions of the message 123456789ABCDEF and key 133457799BBCDFF1. They were performed on a 1.8 Ghz Core2Duo computer running Ubuntu 10.04.
\begin{multicols}{4}
Test 1:  5.960 sec

Test 2:  6.012 sec

Test 3:  5.903 sec

Average: 5.958 sec
\end{multicols}
\begin{verbatim}
  % of   # of      # of
  Time  Seconds    Calls      Name
 42.90    6.10   16,000,000   DES::f()
 24.40    3.47    1,000,000   DES::keyschedule()
 24.19    3.44    1,000,000   DES::algorithm()
  2.04    0.29    1,000,000   DES::inverseInitPermutation()
  2.04    0.29    1,000,000   std::swap_ranges<T,T>()
  1.30    0.18   32,000,000   std::iter_swap<T,T>()
  1.02    0.14   32,000,000   std::__iter_swap<G>::iter_swap<T,T>()
  0.91    0.13    1,000,000   DES::initPermutation()
  0.70    0.10   32,000,000   std::swap<unsigned char>()
  0.21    0.03    1,000,000   DES::~DES()
  0.14    0.02    1,000,000   DES::DES()
  0.07    0.01    1,000,000   DES::encrypt()
  0.07    0.01                main
\end{verbatim}

\section{Revised Runtime Analysis}
  One of the major changes was the f (feistel) function. It went from 23\% to 43\% of the total runtime.  This change is largely because we removed the mixer function. Mixer accounted for 41\% of the runtime, 63\% of mixer's calls were from feistel.  From this we can deduce that 26\% out of the 41\% was directly due to f, and we could say that f really caused 23\% + 26\% = 49\% of the runtime.  From this we can see that the f function's overall contribution to execution time went down without removing any of its duties.

\pagebreak
\section{Developers manual}

@@

... test output ...
\end{verbatim}
}
\section{Users Manual}

\noindent In order to encrypt a file using our implementation the user must run the executable \textbf{encrypt\_opt}.
\begin{verbatim}
  Usage: ./encrypt_opt <infile> <outfile> <key>

    <infile>  - The file you wish to encrypt
    <outfile> - The resulting encrypted file
    <key>     - 8 character string for your key
\end{verbatim}

\noindent \textbf{Ex: ./encrypt\_opt file.jpg file.enc.jpg 12345678} \\

\noindent Similarly to decrypt the file  user has a similar choice for decrypting, both an optimized and original implementation exist.
\begin{verbatim}
  Usage: ./decrypt_opt <infile> <outfile> <key>

    <infile>  - The file you wish to decrypt
    <outfile> - The resulting decrypted file
    <key>     - 8 character string for your key
    <key>     - 8 character string for your key
\end{verbatim}

\noindent \textbf{Ex: ./decrypt\_opt file.enc.jpg file.dec.jpg 12345678} \\

The user should be aware that two set's of of the program exist and are denoted by the \textbf{\_opt} for optimized and \textbf{\_orig} for original appended to each.
These exist so you can easily see the performance difference between the two implementations your self on large files.

@@

  \noindent \textbf{Brian:}
  \begin{itemize}
    \item Unit test setup
    \item Build system
    \item Gathering of all constants and permutation blocks.
    \item Key schedule
    \item Main DES algorithm
  \end{itemize}

 \noindent \textbf{Sam:}
  \begin{itemize}
    \item Bit operators and converters
    \item Feistel function
    \item Implementation API design
    \item Mix function to handle mixing
    \item Block mode implementation
  \end{itemize}

@@

  \bibitem{spec}{ \href{ http://csrc.nist.gov/publications/fips/fips46-3/fips46-3.pdf }{ Data Encryption Standard Specification } }
  \bibitem{deepcrack}{ \href{ http://en.wikipedia.org/wiki/EFF_DES_cracker }{  DES Challenge  } }
  \bibitem{withdrawn}{ \href{ http://csrc.nist.gov/publications/fips/05-9945-DES-Withdrawl.pdf }{ Federal Register document, stating the withdrawl of FIPS-46 } }
  \bibitem{gtest}{ \href{ http://code.google.com/p/googletest/ }{ Google Unit Testing Framework } }
  \bibitem{ wiki  }{ \href{ http://en.wikipedia.org/wiki/Data_Encryption_Standard }{ Wikipedia: Data Encryption Standard } }
  \bibitem{ block }{ \href{ http://en.wikipedia.org/wiki/Block_cipher_modes_of_operation }{ Wikipedia: Block Cipher Modes } }
  \bibitem{ desge }{ \href{ http://www.comms.scitech.sussex.ac.uk/fft/crypto/des_algorithm_details.txt}{ How to implement the Data Encryption Standard (DES) } }
  \bibitem{ desg2 }{ \href{ http://orlingrabbe.com/des.htm }{ DES Illustrated } }
  \bibitem{ gprof }{ \href{ http://linux.die.net/man/1/gprof }{ GProf Manual Page } }
  \bibitem{ g++ }{ \href{ http://linux.die.net/man/1/g++ }{ GCC Manual Page } }
